---
title: "ciftiTools Demo"
date: "5/25/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Getting started

#### Preliminaries

Install and load `ciftiTools`:

```{r}
if (!require("ciftiTools", quietly=TRUE)){ 
  install.packages("ciftiTools")
} else if (packageVersion("ciftiTools") < "0.9.0") {
  update.packages("ciftiTools")
}

library(ciftiTools)
```

Download the Connectome Workbench here: https://www.humanconnectome.org/software/get-connectome-workbench

Tell `ciftiTools` where to find the Connectome Workbench:

```{r}
# Replace '~/Desktop/workbench' with the actual path on your computer.
#   If successful, the path to the Workbench executable will be printed.
ciftiTools.setOption("wb_path", "~/Desktop/workbench")
```

Setup `rgl` plots for knitting.

```{r}
library(rgl)
rgl::setupKnitr()

# Sometimes the first OpenGL window does not render properly.
rgl::rgl.open(); rgl::rgl.close()
```

#### First steps

Read in our first CIFTI file!

```{r}
cii_fname <- file.path("data", "MyelinAndCorrThickness.dscalar.nii")
xii <- read_cifti(cii_fname)
xii # same as `summary(xii)`
```

Plot it:

```{r, fig.cap="Myelin", rgl=TRUE, format="png", fig.height=3, fig.width=4}
# same as view_xifti(xii) and view_xifti_surface(xii)
plot(xii, zlim=c(1,2)) 
```

View a list of commonly-used functions:

```{r}
if(interactive()) { help(ciftiTools) }
```

View details for a specific function:

```{r}
if(interactive()) { help(view_xifti_surface) }
```

## Task #1

#### a. Read & plot the fMRI data.

Use `read_cifti` to read in the CIFTI file named `cii_fname`. Set the `brainstructures` argument to include all brain structures.

```{r}
cii_fname <- file.path("data/toy_fMRI.dtseries.nii")
# [Use `read_cifti` here]
#xii <- 
```

We will add very inflated surfaces prior to plotting.

```{r}
xii <- add_surf(
  xii, 
  load_surf("left", "very inflated"), 
  load_surf("right", "very inflated")
)
```

Use `view_xifti_surface` to plot the first timepoint.

Customize your plot by specifying some arguments! Here's a few ideas:

* Set `title` 
* Set `zlim` (min, max)
* Change `colors` to the name of a sequential palette from `RColorBrewer`. To see a list, use `RColorBrewer::display.brewer.all()`.

```{r, fig.cap="First timepoint", rgl=TRUE, format="png", fig.height=3, fig.width=4}
# [Use `view_xifti_surface` here]
# view_xifti_surface(
```

####  b. Compute & plot the fALFF

Here's a function to calculate the fALFF of a single time series vector. 

```{r}
#' Adapted from `ANTsR::alffmap`
falff <- function(x, TR=.72, f_low=.01, f_high=.1){
  # Subtract linear trend
  x <- lm(x~seq(length(x)))$residuals
  
  # Compute Fourier decomposition.
  y <- stats::spec.pgram(
    stats::ts(x, frequency=1/TR), 
    taper=0, plot=FALSE
  )
  freq_mask <- (y$freq > f_low) & (y$freq < f_high)
  
  # Compute proportion within frequency range
  sum(y$spec[freq_mask]) / sum(y$spec)
}
```

Use `apply_xifti` to apply the `falff` function to each row of the data matrix in `xii`.

```{r}
# [Use `apply_xifti` here.]
# fxii <- 
```

Plot the results using a color scale going from 0 (white) to 1 (blue), and save to a PNG file. For help on finding the relevant arguments, read `help(view_xifti_surface)`.

```{r fALFF, fig.cap="fALFF", rgl=TRUE, format="png", fig.height=3, fig.width=4}
# [Use `view_xifti_surface` here.]
# view_xifti_surface(
```

```{r fALFFsub, fig.cap="fALFF, subcortex", format="png", fig.height=4, fig.width=2.5}
# [Use `view_xifti_volume` here.]
# view_xifti_volume(
```

## Task #2

Read the fMRI data and parcellation.

```{r}
cii_fname <- file.path("data/MSC03_rest07.4k.dtseries.nii")
xii <- read_cifti(cii_fname)
dim(xii)
```

```{r}
parc <- read_cifti("data/Kong2022_100Parc.dlabel.nii", resamp_res=4000)
dim(parc)
```

```{r}
# rownames(parc$meta$cifti$labels[[1]])
table(c(as.matrix(parc))==0)
```

Incompatible ROIs: we need to replace the medial wall mask with `NA` values.

```{r}
xii <- move_from_mwall(xii)
```

Pick our seed and plot it.

```{r parc, fig.cap='Parcellation', rgl=TRUE, fig.width=4, fig.height=3}
# View(parc$meta$cifti$labels$parcels["Key"])
plot(parc, title="Parcellation")
```

```{r parc2, fig.cap='Selected parcel', rgl=TRUE, fig.width=4, fig.height=3}
the_seed <- 3 # sample(seq(100), 1) 
parc_masked <- transform_xifti(parc, function(q){ifelse(q==the_seed, q, 0)})
plot(parc_masked, borders="black", title=paste0("Parcel #", the_seed), fname="Seed")
```

Now that we've identified the seed, let's do this!

#### a. Scrub the fMRI data.

Use `select_xifti` to mask out high-motion timepoints. (Only keep the columns for which the corresponding entry of `xii_tmask` is `TRUE`.)

```{r}
xii_tmask <- read.csv("data/MSC03_rest07_tmask.txt")
xii_tmask <- which(as.logical(xii_tmask[,1]))
# [Use `select_xifti` here.] 
# xii <- select_xifti(
dim(xii)
```

#### b. Compute the seed correlations.

The below code chunk is already filled out. Here's what it does:

* `seed_mask`: Convert the parcellation `"xifti"` to a numeric vector of parcel indices for each vertex, and then convert that to a binary vector indicating whether the parcel is the seed parcel.
* `parc_mean_ts`: Convert the resting-state fMRI `"xifti"` to a matrix, subset the rows to select only vertices inside the parcel, and then take the mean across space.
* `xii_sc`: Compute correlation between each vertex time series, and the parcel mean time series. 

```{r}
seed_mask <- c(as.matrix(parc)) == the_seed
parc_mean_ts <- colMeans(as.matrix(xii)[seed_mask,,drop=FALSE])
xii_sc <- cor(t(as.matrix(xii)), parc_mean_ts)
xii_sc <- newdata_xifti(xii, xii_sc)
```

Now we will convert the result back to a `"xifti"` with `newdata_xifti`, and then format is as a "dscalar" since the data no longer represents a time series.

```{r}
xii_sc <- cor(t(as.matrix(xii)), parc_mean_ts)
# [Use `newdata_xifti` here.]
# xii_sc <- newdata_xifti(
# [Use `convert_xifti` here.]
# xii_sc <- convert_xifti(
```

Write out the result with `write_xifti`.

```{r}
seed_fname <- paste0("seed_cor_", the_seed, ".dscalar.nii")
# [Use `write_xifti` here.]
# write_xifti(
```

Finally, plot the result!

```{r seedCor, fig.cap='Seed correlation', rgl=TRUE, fig.width=4, fig.height=3}
view_xifti_surface(xii_sc, zlim=c(-.7, .7))
```

```{r seedCorSub, fig.cap='Seed cor., subcortex', fig.width=2.5, fig.height=4}
view_xifti_volume(xii_sc, zlim=c(-.7, .7))
```


## Task #3

Install the latest version of BayesfMRI from GitHub.

```{r}
if (!requireNamespace("BayesfMRI", quietly=TRUE) || packageVersion("BayesfMRI") < "0.2.0") {
  if (!requireNamespace("devtools", quietly=TRUE)) { install.packages("devtools") }
  devtools::install_github("mandymejia/BayesfMRI", "2.0")
}
library(BayesfMRI)
```

Read in and format the event onsets file. Use the "left foot" task.

```{r}
onsets <- paste0("data/MSC03_motor07_run0", seq(2), "_events.tsv")
onsets <- lapply(as.list(onsets), read.csv, sep="\t")
onsets <- lapply(onsets, function(q){list(as.matrix(q[q[,3]=="LFoot",seq(2)]))})
```

Run classical GLM for the right and left cortex separately.

```{r}
bglmR <- BayesGLM_cifti(
  paste0("data/MSC03_motor07_run0", seq(2), ".4k.dtseries.nii"),
  brainstructures="right", surfR = load_surf("right", resamp_res=4000),
  onsets=onsets, ar_order=0, EM=FALSE, Bayes=FALSE, resamp_res=NULL, TR=2.2
)

bglmL <- BayesGLM_cifti(
  paste0("data/MSC03_motor07_run0", seq(2), ".4k.dtseries.nii"),
  brainstructures="left", surfL = load_surf("left", resamp_res=4000),
  onsets=onsets, ar_order=0, EM=FALSE, Bayes=FALSE, resamp_res=NULL, TR=2.2
)
```

#### a. Write `combine_and_plot_GLM`

This function should combine classical, multi-session GLM results from the left and right hemispheres, and then plot the result. To make it work, add a line that uses `combine_xifti` to merge the left and right hemisphere `"xifti"` objects.

```{r}
combine_and_plot_GLM <- function(
  left_GLM, right_GLM, 
  what=c("estimates", "SE_estimates", "resids"),
  ...){
  
  # Argument checks.
  stopifnot(is.xifti(left_GLM$betas_classical$session_avg))
  stopifnot(is.xifti(right_GLM$betas_classical$session_avg))
  what <- match.arg(what, c("estimates", "SE_estimates", "resids"))
  
  # Get separate `"xifti"` objects.
  if (what=="estimates"){
    left_GLM <- left_GLM$betas_classical$session_avg
    right_GLM <- right_GLM$betas_classical$session_avg
    
  } else if (what=="SE_estimates") {
    # Left
    left_mask <- left_GLM$GLMs_classical$cortexL$session_avg$mask
    x <- left_GLM$GLMs_classical$cortexL$session_avg$SE_estimates
    left_GLM <- newdata_xifti(
      left_GLM$betas_classical$session_avg,
      x[left_mask,,drop=FALSE]
    )
    # Right
    right_mask <- right_GLM$GLMs_classical$cortexR$session_avg$mask
    x <- right_GLM$GLMs_classical$cortexR$session_avg$SE_estimates
    right_GLM <- newdata_xifti(
      right_GLM$betas_classical$session_avg,
      x[right_mask,,drop=FALSE]
    )
    
  } else if (what=="resids") {
    left_GLM <- newdata_xifti(
      left_GLM$betas_classical$session_avg,
      left_GLM$GLMs_classical$cortexL$session_avg$resids
    )
    right_GLM <- newdata_xifti(
      right_GLM$betas_classical$session_avg,
      right_GLM$GLMs_classical$cortexR$session_avg$resids
    )
  } else { stop() }
  
  # [Use `combine_xifti` here.]
  # both_GLM <- combine_xifti(
  
  plot(both_GLM, ...)
}
```

Try it out!

```{r bglm, fig.cap='Combined GLM results', fig.width=8, fig.height=3}
combine_and_plot_GLM(bglmL, bglmR, "SE")
```

Lastly, let's check out an example of using `view_comp`. We will plot the GLM result alongside its activations calculated by statistical inference.

```{r}
# Write and save the GLM result plot.
p1 <- plot(bglmR, fname="glmR_vals", title="GLM result", together="leg")
# Write and save the activations plot.
p2 <- plot(id_activations_cifti(bglmR)$activations_xifti, fname="glmR_act", title="Thresholded activation", together="leg")
```

```{r bglmR, fig.cap='Left foot task activation (right cortex)', fig.width=7}
# Composite
view_comp(c(p1, p2), nrow=1)
```

## Appendix

#### Making plots for publications

`ciftiTools` can create high-quality plot suitable for publication. Here are a few tips:

* Save PNG files using `fname` rather than taking screenshots of the OpenGL window.
* Use `width` and `height` to increase the plot size. The limits are equal to your screen's resolution.
* `view_comp` can create composite images, but it's also possible to put together plots by importing them into an image-editing software like GIMP, or to create subplots within `LaTeX`. For this same reason, you may want to omit `title` and instead add text above or below the plot later. You can also use `legend_embed=FALSE` to composite the legend yourself. 
* When plotting `dlabel` files where colors are repeated for different keys, you can use `borders=TRUE` to delineate them better on cortex plots. 
* In Photoshop, the "Object Selection" tool works great with the brain meshes! You can also try using the `material=list(smooth=FALSE, lit=FALSE)` to avoid smoothing colors and shading, allowing precise selection of region by color/value for further editing. 

Please see the Appendix C in the [`ciftiTools` paper](https://doi.org/10.1016/j.neuroimage.2022.118877) for guidance on embedding `ciftiTools` plots in R Markdown documents.